---
title: "Primer design"
author: "Alexey Uvarovskii"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
  max-width: 900px;
}
</style>

## Primer design for transcripts and arbitrary sequences

The package allows to run the Primer3 with user specified parameters.
There is a shortcut for design, in which one of the primers
must overlap with the splice junction.

### A quick start

```{r eval = FALSE}
library(primex)
exon1 <- paste0(
  "CTCACCATGGATGATGATATCGCCGCGCTCGTCGTCGACAACGGCTCCGGCATGTGCAAG",
  "GCCGGCTTCGCGGGCGACGATGCCCCCCGGGCCGTCTTCCCCTCCATCGTGGGGCGCCCC",
  "AGGCACCAG")
exon2 <- paste0(
 "GGCGTGATGGTGGGCATGGGTCAGAAGGATTCCTATGTGGGCGACGAGGCCCAGAGCAAG",
 "AGAGGCATCCTCACCCTGAAGTACCCCATCGAGCACGGCATCGTCACCAACTGGGACGAC",
 "ATGGAGAAAATCTGGCACCACACCTTCTACAATGAGCTGCGTGTGGCTCCCGAGGAGCAC",
 "CCCGTGCTGCTGACCGAGGCCCCCCTGAACCCCAAGGCCAACCGCGAGAAGATGACCCAG") 

seqOpts <- seqSettings(
  seqId = "example1",
  seq = c(exon1, exon2))

primers <- runPrimer3(seqOpts)
```

### A quick start with annotations

```{r eval = FALSE}
library(primex)  

library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.NCBI.GRCh38)
db <- EnsDb.Hsapiens.v86
bsg <- BSgenome.Hsapiens.NCBI.GRCh38

# define sequences
geneName <- "BCL6"
exonsByTx <- ensembldb::exonsBy(db, filter = GenenameFilter(geneName))
# find unique splice junctions 
sjExonPairs <- selectPairs(exonsByTx)
# or report all the pairs
# sjExonPairs <- selectPairs(exonsByTx, tolerance = -1)

# take your transcript of interest
exseqs <- addSeq(sjExonPairs$ENST00000419510, bsg)

# run Primer3 for one of splice junction pair
seqOpts <- seqSettings(gr = exseqs$`ENSE00001666929|ENSE00001377234`)
primers <- runPrimer3(seqOpts)
```

## Step by step

```{r}
library(primex)  
```

### Find unique exons 

Let us consider human BCL6 gene, which has a lot of isoforms.
First, we need to get annotation e.g. from Ensembl data base.
The input format is a list of genomic ranges (`GRanges` object).
Every list item has corresponding transcript name and a list 
of exons with their `exon_id` and `exon_rank`.

```{r fig.width=6, fig.height=4, dpi = 300, out.width="70%"}
suppressPackageStartupMessages(
  library(GenomicRanges)
)
primex::bcl6exons[1]
plotSegments(primex::bcl6exons)
```


In order to distinguish isoforms using PCR, we look for unique splice junctions,
i.e. unique exon pairs. **This is the default behaviour. **
For every transcript we may receive several such 
candidates, for example,
```{r}
sjExonPairs <- selectPairs(primex::bcl6exons)
sjExonPairs$ENST00000419510[1]
```

### Provide sequences 

If there is annotation package, one may retrieve sequences of the exons from it.
We implemented a function to tie sequences from the  BSgenome packages:

```{r}
suppressPackageStartupMessages(
  library(BSgenome.Hsapiens.NCBI.GRCh38)
)
bsg <- BSgenome.Hsapiens.NCBI.GRCh38
exseqs <- addSeq(sjExonPairs$ENST00000419510, bsg)
exseqs$`ENSE00001666929|ENSE00001377234`
```

### Run Primer3

The *Primer3* program has a lot of parameters.
There are two sets, *SEQUENCE* parameters and *PRIMER* parameters.
The *SEQUENCE* parameters describe such items, as a sequence identificator,
the sequence of nucleotides itself, 
splice junction position, etc. One may start from the `GRanges` object, 
which has a column `seq` with a nucleotide sequence for every item:
```{r}
seqOpts <- seqSettings(gr = exseqs$`ENSE00001666929|ENSE00001377234`)
res <- runPrimer3(seqOpts)
# take first two pairs: Primer3 reports by default maximum 5 pairs
knitr::kable(t(res$primers)[,1:2])
```

### More on parameters

All the parameters can be explicitly specified:
```{r}
seqOpts <- seqSettings(seqId = "example2", seq = c("AAATGCTGAAGGT"))
seqOpts$SEQUENCE_TARGET <- "37,21"
str(seqOpts)
```
The parameters must be set as character strings.  

The run settings are set using `p3Settings`:
```{r}
primerOpts <- p3Settings()
primerOpts$PRIMER_PRODUCT_OPT_SIZE <- "100,300"
primerOpts$PRIMER_OPT_TM <- "60.0"
```

It is possible to run your own *Primer3* executable, having
provided the needed paths.
```{r eval = FALSE}
primerOpts <- p3Settings(
  defaultsFile = 
    "/home/tsawyer/primer3/primer3_v1_1_4_default_settings.txt")
runPrimer3(
  seqOpts,
  primerOpts,
  returnStats = TRUE,
  path = list(primer3 = "/home/tsawyer/primer3/primer3_core",
              config  = "/home/tsawyer/primer3/primer3_config")
)
```

Please, refer to the [Primer3 manual page](http://primer3.sourceforge.net/primer3_manual.htm) for more details.

## No primers?

Let us have two subsequent exons and we would like to design a pair of 
primers for them.
```{r}
exon1 <- paste0(
  "CTCACCATGGATGATGATATCGCCGCGCTCGTCGTCGACAACGGCTCCGGCATGTGCAAG",
  "GCCGGCTTCGCGGGCGACGATGCCCCCCGGGCCGTCTTCCCCTCCATCGTGGGGCGCCCC",
  "AGGCACCAG")
exon2 <- paste0(
 "GGCGTGATGGTGGGCATGGGTCAGAAGGATTCCTATGTGGGCGACGAGGCCCAGAGCAAG",
 "AGAGGCATCCTCACCCTGAAGTACCCCATCGAGCACGGCATCGTCACCAACTGGGACGAC",
 "ATGGAGAAAATCTGGCACCACACCTTCTACAATGAGCTGCGTGTGGCTCCCGAGGAGCAC",
 "CCCGTGCTGCTGACCGAGGCCCCCCTGAACCCCAAGGCCAACCGCGAGAAGATGACCCAG") 

```

It is straightforward to call Primer3 with the default values:
```{r}
seqOpts <- setSeqOptions(
  seqId = "example1",
  seq = c(exon1, exon2))

result <- runPrimer3(seqOpts)
result$primers
```

We can see, that there are no primers returned. Let us see, 
what can be a reason for their rejection:
```{r}
diagnose(result)
```

By default, `runPrimer3` overwrites the option 
`PRIMER_EXPLAIN_FLAG`, which allows to see the run statistics.

